Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method

### 2023-08-07 manu-draft.qmd
fixing units in tabs and reproc misdat
text jusification for priors updated following meeting

```{=tex}
\newcommand{\auni}{\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)}
\newcommand{\runi}{\text{mmol}/\text{m}^3/\text{d}}
\newcommand{\buni}{\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)}
\newcommand{\area}{\text{mmol}/\text{m}^2/\text{d}}
```
## Abstract {.unnumbered}

@@

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each. As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches. Reasonable uninformed prior distributions for EBASE were chosen with mean values using approximate estimates from the literature [@caffrey2004; @Wanninkhof14] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space. Specifically, the prior mean for $R$ was chosen as 300 $\text{mmol}/\text{m}^2/\text{d}$ (\~10 $\text{g O}_2/\text{m}^2/\text{d}$) using a national synthesis of metabolic rates for US estuaries [Table 2 in @caffrey2004]. Although @caffrey2004 focused on small, shallow estuaries that are part of the NERRS, this dataset is the largest, longest, and most consistently collected in the US and it is reasonable to expect others may apply EBASE to these data. The prior mean for $a$ was chosen assuming that $P$ and $R$ are comparable at long time scales, such that $aPAR$ = $R$, or $a$ = $300/PAR$. Using the global mean surface $PAR$ of 70 $W/m^2$, the prior mean for $a$ is reasonably set at 4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. Lastly, the prior mean for $b$ was chosen as 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following @Wanninkhof14. For the prior standard deviations.... Together, the default priors for $a$, $R$, and $b$ are:

@@

```{r}
wqmis <- misdat$Temp # same percentages across all wq parms
mtmis <- misdat$totpar # same percentages across all met parms
```

@@

The prior distributions for the EBASE parameters follow truncated normal distributions with defined means ($\mu$) and standard deviations ($\sigma$)(Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the priors. For each parameter, low and high values for the means and standard deviations were assessed as changes approximately equal to two orders of magnitude of the means and standard deviations of the known parameters of the synthetic time series. The only exception was the $b$ parameter where the low values created priors with a distribution close to zero and the high values were equivalent to a uniform prior distribution in the range of 0 to 0.502 (justification noted in @sec-ebasemod). Each unique combination of low and high values for the means and standard deviations of each parameter were evaluated, creating 2$^6$ = 64 different combinations of prior distributions that were used with the synthetic time series in EBASE.

The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $R$, and $b$). Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being one day and a maximum being the length of the time series. For a chosen optimization period, single estimates for $a$, $R$, and $b$ are returned by the model. A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if variation occurs at time scales less than the optimization period. Results from optimization periods of one, seven, and thirty days were compared for the synthetic one-year time series. Each optimization period was also evaluated for every unique combination of prior distributions, totaling 192 (64 times the three optimization periods) unique analyses for the assessment of changing priors of all parameters.

>>>>

```{=tex}
\newcommand{\auni}{\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)}
\newcommand{\runi}{\text{mmol}/\text{m}^2/\text{d}}
\newcommand{\buni}{\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)}
```
## Abstract {.unnumbered}

@@

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each. As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches. Reasonable uninformed prior distributions for EBASE were chosen with mean values using approximate estimates from the literature [@caffrey2004; @Wanninkhof14] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space. Specifically, the prior mean for $R$ was chosen as 300 $\text{mmol}/\text{m}^2/\text{d}$ (\~10 $\text{g O}_2/\text{m}^2/\text{d}$) using a national synthesis of metabolic rates for US estuaries [Table 2 in @caffrey2004]. Although @caffrey2004 focused on small, shallow estuaries that are part of the NERRS, this dataset is the largest, longest, and most consistently collected in the US and it is reasonable to expect others may apply EBASE to these data. The prior mean for $a$ was chosen assuming that $P$ and $R$ are comparable at long time scales, such that $aPAR$ = $R$, or $a$ = $300/PAR$. Using the global mean surface $PAR$ of 70 $W/m^2$, the prior mean for $a$ is reasonably set at 4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. Lastly, the prior mean for $b$ was chosen as 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following @Wanninkhof14. The prior standard deviations were chosen similarly based on values in @caffrey2004 based on the standard deviation of the reported values. The standard deviation of $R$ was 171 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ and the standard deviation for $a$ was set proportionally as 2.25 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, again assuming $P$ = $aPAR$ = $R$. The standard deviation for $b$ was set as 50% of the mean as an uninformed and reasonable value given that there are no comparable global or national estimates for the range of values. Together, the default priors for $a$, $R$, and $b$ are:

@@

```{r}
wqmis <- misdat$Temp # same percentages across all wq parms
mtmis <- misdat$PAR # same percentages across all met parms
```

@@

The prior distributions for the EBASE parameters follow truncated normal distributions with defined means ($\mu$) and standard deviations ($\sigma$)(Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the priors. Low and high values for the means and standard deviations were assessed following reasonable ranges from @caffrey2004 using similar logic for the default priors. The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ and proportionally for $a$ as 2.0 and 14.4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default mean values to test a range from high precision to completely uninformed values. The mean and standard deviation for the $b$ parameter were not tested and set as the default values for all analyses given that there is minimal prior knowledge on expectations for this parameter. The default prior distribution is already uninformed and we have no appropriate expectations on methods for varying this parameter. Each unique combination of low and high values for the means and standard deviations of each parameter were evaluated, creating 2$^4$ = 16 different combinations of prior distributions that were used with the synthetic time series in EBASE.

The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $R$, and $b$). Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being one day and a maximum being the length of the time series. For a chosen optimization period, single estimates for $a$, $R$, and $b$ are returned by the model. A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if variation occurs at time scales less than the optimization period. Results from optimization periods of one, seven, and thirty days were compared for the synthetic one-year time series. Each optimization period was also evaluated for every unique combination of prior distributions, totaling 48 (16 times the three optimization periods) unique analyses for the assessment of changing priors of all parameters.
